Universals of word order result from optimization of grammars for efficient communication

### 2019-01-20 writeup/paper-names.tex
revising paper

\begin{dependency}[theme=simple]
    \begin{deptext}
    I \& wrote \& a \& long \& letter \& to \& a \& friend \\
    \end{deptext}
    \depedge{2}{1}{nsubj}
    \depedge{2}{5}{obj}
    \depedge{5}{4}{ }
    \depedge{5}{3}{ }
    \depedge{2}{6}{}
    \depedge{6}{8}{ }
    \depedge{8}{7}{ }
\end{dependency}
	\caption{A sentence in Arabic (top) and Japanese (bottom), translating to `I wrote a letter to a friend.' Note the reversal of word order: Arabic has verb-object order and prepositions, while Japanese has object-verb order and postpositions.}\label{fig:arabic-japanese}
\end{figure}

@@

Let $t$ be an unordered dependency tree, i.e. a structure as in Figure~\ref{fig:arabic-japanese} but without any specified ordering of the words. We think of an unordered dependency tree $t$ as representing the compositional meaning to be conveyed by a sentence. We define a \key{word order grammar} as a function $L_\theta(t)$, whose behavior is specified by parameters $\theta$, which takes an unordered dependency tree $t$ as input and produces as output a probability distribution over ordered sequences of words $\mathbf{w}$ linearizing the tree, or equivalently, a probability distribution over orders on the nodes of the tree. % do we need to define "linearize"?

%For example, TODO % We should have some pretty picture visualizing the process here

@@

@@

See the SI for further details on the joint optimization of $\theta$ and $\psi$ and on the neural parser architecture.

\paragraph{Efficiency}
The two scoring functions---predictability and parseability---can be combined into an overall equation of the form:
\begin{equation}
\label{eq:combined-score}
R_{Efficiency} = R_{Pars} + \lambda R_{Pred},
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$. This combined equation represents efficiency: bits of information transferred about dependency trees minus the cost of the transmission. In expectation over utterances (substituting Eq.~\ref{eq:combined-score} into Eq.~\ref{eq:exp-over-utts}), the overall objective function to be maximized by a word order grammar can be written in information-theoretic terms as:
\begin{equation}
    \label{eq:efficiency}
    J_T(\theta) = I[L_\theta(T);T] - \lambda H[L_\theta(T)],
\end{equation}
where $I[L_\theta(T);T]$ is the \key{mutual information} between strings and trees, and $H[L_\theta(T)]$ is the \key{entropy} of strings \cite{cover2006elements}. Maximizing Eq.~\ref{eq:efficiency} is equivalent to maximizing mutual information between strings and trees subject to a constraint on the maximum entropy of strings.

The two terms of Eq.~\ref{eq:efficiency} implement two opposing pressures. Parseability (the mutual information term) pushes to make strings more heterogeneous, so that they can indicate different tree structures unambiguously. Predictability (the entropy term) pushes to make strings more homogeneous, to reduce the complexity of using the language. The idea that natural language arises from a tension between these two forces goes back over a century \cite{gabelentz1901sprachwissenschaft}. Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking unordered dependency trees $T$ as the underlying meanings to be conveyed. Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI for the precise relationship).

@@

\begin{figure}
    \centering
    \includegraphics[scale=.4]{figures/pareto-plane-iso-best-balanced-legend.pdf}
    \caption{Predictability and parseability of 51 UD languages (red), indicated by ISO codes, compared to ten baseline word order grammars per language (green). Predictability and parseability scores are $z$-scored within language. Each point for a real language has a line pointing in the direction of the center of mass of its baselines. The green contour shows the density of baseline languages. Unlabeled dots represent the centroid for real languages (red), baseline languages (green), and languages optimized for predictability (yellow), parseability (pink), efficiency (blue), and dependency length (red). When a language is to the bottom--left of its baselines, this indicates that it is relatively optimal for efficiency.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}

@@

\begin{figure*}
\includegraphics[width=0.98\textwidth]{figures/grid-optimized.pdf} % need MLE langs and Efficiency langs on here
\caption{Predictability (top) and parseability (bottom) of real, optimized, and random languages, for English (left) and Japanese (right). We provide cross-entropy for predictability (lower is better) and unlabeled attachment scores---a standard metric from natural language processing---for parseability (higher is better).}\label{fig:predict-pars}
\end{figure*}

@@

\begin{figure}
    \centering
    \includegraphics[scale=.2]{figures/posterior-satisfied-universals-efficiency.pdf}
     \includegraphics[scale=.2]{figures/posterior-satisfied-universals-langmod.pdf}
     \includegraphics[scale=.2]{figures/posterior-satisfied-universals-parseability.pdf}
    \includegraphics[scale=.2]{figures/posterior-satisfied-universals-depl.pdf}
	\caption{Posterior of the number of correlations predicted in the direction found in the UD treebanks.}
    \label{fig:posterior-satisfied-count}
\end{figure}

@@

While the efficiency objective succeeds in predicting the presence of the most word order correlations, the dependency length objective seems to predict a greater prevalence for those correlations that it does predict. We believe this discrepancy is a result of noise in the optimization process. Optimizing for efficiency is computationally challenging, requiring stochastic gradient descent on word order grammar, language model, and parser parameters, while optimizing for dependency length is simpler and thus more optimal grammars can be reached easily. Inasmuch as dependency length minimization is computationally easier to implement than efficiency maximization, we believe it will remain useful as an easy-to-use heuristic for predicting and explaining word order universals.
% TODO we can also report (in the SI) prevalences among the best-performing of the optimized languages, i.e., those where optimization was most successful. I would expect that this should narrow the gap between DepL and Efficiency, which would confirm that the gap is really about the difficulty of optimization, not about the objective functions.

@@

We found that a large subset of the Greenbergian word order correlations can be explained in terms of from optimization of grammars for efficient communication, as defined by information theoretic criteria and implemented using state-of-the-art machine learning methods. We defined a space of word order grammars, as well as objective functions reflecting communicative efficiency, and found that the word order grammars that maximize communicative efficiency reproduce the word order universals. Beyond our present results, we provide a complete formalization and computational framework in which theories of the functional optimization of languages can be tested. Other objective functions could be hypothesized and tested within our framework; furthermore, future advances in machine learning will enable the optimization of richer and richer models of grammar.

>>>>

\begin{dependency}[theme=simple]
    \begin{deptext}
    I \& wrote \& a \& letter \& to \& a \& friend \\
    \end{deptext}
    \depedge{2}{1}{nsubj}
    \depedge{2}{4}{obj}
    \depedge{4}{3}{ }
    \depedge{4}{3}{ }
    \depedge{2}{5}{}
    \depedge{5}{7}{ }
    \depedge{7}{6}{ }
\end{dependency}
	\caption{A sentence in Arabic (top) and Japanese (bottom), translating to `I wrote a letter to a friend.' Note the reversal of word order: Arabic has verb-object order and prepositions, while Japanese has object-verb order and postpositions.}\label{fig:arabic-japanese}
\end{figure}

@@

Let $t$ be an unordered dependency tree, i.e. a structure as in Figure~\ref{fig:arabic-japanese} but without any specified ordering of the words. We think of an unordered dependency tree $t$ as representing the compositional meaning to be conveyed by a sentence. We define a \key{word order grammar} as a function $L_\theta(t)$, whose behavior is specified by parameters $\theta$, which takes an unordered dependency tree $t$ as input and produces as output a probability distribution over ordered sequences of words $\mathbf{w}$ linearizing the tree, or equivalently, a probability distribution over orders on the nodes of the tree. % do we need to define "linearize"?
This process is visualized in Figure~\ref{fig:process-visualization}.

@@

\begin{figure}
\includegraphics[width=0.48\textwidth]{figures/process-visualization.png}
\caption{Our method for constructing counterfactual versions of languages: We extract unordered dependency tree topologies from dependency corpora, and apply word order grammars to obtain counterfactually reordered corpora.}\label{fig:process-visualization}
\end{figure}

@@

See the SI (appendix section 6-7) for further details on the joint optimization of $\theta$ and $\psi$ and on the neural parser architecture, and see SI (section 8) for robustness to different parsing models.

\paragraph{Efficiency}
The two scoring functions---predictability and parseability---can be combined into an overall equation of the form:
\begin{equation}
\label{eq:combined-score}
R_{Efficiency} = R_{Pars} + \lambda R_{Pred},
\end{equation}
with an interpolation weight $\lambda \in [0,1)$. In all experiments in this paper we use $\lambda=.9$ (see SI appendix section 5 for mathematical justification). This combined equation represents efficiency: bits of information transferred about dependency trees minus the cost of the transmission. In expectation over utterances (substituting Eq.~\ref{eq:combined-score} into Eq.~\ref{eq:exp-over-utts}), the overall objective function to be maximized by a word order grammar can be written in information-theoretic terms as:
\begin{equation}
    \label{eq:efficiency}
    J_T(\theta) = I[L_\theta(T);T] - \lambda H[L_\theta(T)],
\end{equation}
where $I[L_\theta(T);T]$ is the \key{mutual information} between strings and trees, and $H[L_\theta(T)]$ is the \key{entropy} of strings \cite{cover2006elements}. Maximizing Eq.~\ref{eq:efficiency} is equivalent to maximizing mutual information between strings and trees subject to a constraint on the maximum entropy of strings.

The two terms of Eq.~\ref{eq:efficiency} implement two opposing pressures. Parseability (the mutual information term) pushes to make strings more heterogeneous, so that they can indicate different tree structures unambiguously. Predictability (the entropy term) pushes to make strings more homogeneous, to reduce the complexity of using the language. The idea that natural language arises from a tension between these two forces goes back over a century \cite{gabelentz1901sprachwissenschaft}. Eq.~\ref{eq:efficiency} is a special case of the objective function proposed in \cite{ferrericancho2002zipf,ferrericancho2007global,futrell2017memory} as a general objective for communicative systems, taking unordered dependency trees $T$ as the underlying meanings to be conveyed. Eq.~\ref{eq:efficiency} can also be seen as a simplified form of the Information Bottleneck \cite{tishby1999information}, a general objective function for lossy compression which has recently been applied to explain linguistic phenomena such as color naming systems \cite{zaslavsky2018efficient} (see SI section 4 for the precise relationship).

@@

\begin{figure}
    \centering
    \includegraphics[scale=.45]{../results/plane/pareto-plane-iso-best-balanced-legend.pdf}
    \caption{Predictability and parseability of 51 UD languages (red), indicated by ISO codes, compared to ten baseline word order grammars per language (green). Predictability and parseability scores are $z$-scored within language. Each point for a real language has a line pointing in the direction of the center of mass of its baselines. The green contour shows the density of baseline languages. Unlabeled dots represent the centroid for real languages (red), baseline languages (green), and languages optimized for predictability (yellow), parseability (pink), efficiency (blue), and dependency length (red). When a language is to the bottom--left of its baselines, this indicates that it is relatively optimal for efficiency.} % TODO colors, fontsize, etc could conceivably be made prettier
    \label{fig:pareto-plane}
\end{figure}

@@

\begin{figure*}
\includegraphics[width=0.98\textwidth]{../results/plot-by-lang/grid-optimized.pdf} % need MLE langs and Efficiency langs on here
\caption{Predictability (top) and parseability (bottom) of real, optimized, and random languages, for English (left) and Japanese (right). We provide cross-entropy for predictability (lower is better) and unlabeled attachment scores---a standard metric from natural language processing---for parseability (higher is better).}\label{fig:predict-pars}
\end{figure*}

@@

\begin{figure}
    \centering
     \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-efficiency.pdf}
     \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-langmod.pdf}
     \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-parseability.pdf}
     \includegraphics[scale=.2]{../results/correlations/figures/posterior-satisfied-universals-depl.pdf}
	\caption{Posterior of the number of correlations predicted in the direction found in the UD treebanks.}
    \label{fig:posterior-satisfied-count}
\end{figure}

@@

While the efficiency objective succeeds in predicting the presence of the most word order correlations, the dependency length objective seems to predict a greater prevalence for those correlations that it does predict. We believe this discrepancy is a result of noise in the optimization process. Optimizing for efficiency is computationally challenging, requiring stochastic gradient descent on word order grammar, language model, and parser parameters, while optimizing for dependency length is simpler and thus more optimal grammars can be reached easily (See SI section 2 for supporting data). Inasmuch as dependency length minimization is computationally easier to implement than efficiency maximization, we believe it will remain useful as an easy-to-use heuristic for predicting and explaining word order universals.
% TODO we can also report (in the SI) prevalences among the best-performing of the optimized languages, i.e., those where optimization was most successful. I would expect that this should narrow the gap between DepL and Efficiency, which would confirm that the gap is really about the difficulty of optimization, not about the objective functions.

@@

We found that a large subset of the Greenbergian word order correlations can be explained in terms of optimization of grammars for efficient communication, as defined by information theoretic criteria and implemented using state-of-the-art machine learning methods. We defined a space of word order grammars, as well as objective functions reflecting communicative efficiency, and found that the word order grammars that maximize communicative efficiency reproduce the word order universals. Beyond our present results, we provide a complete formalization and computational framework in which theories of the functional optimization of languages can be tested. Other objective functions could be hypothesized and tested within our framework; furthermore, future advances in machine learning will enable the optimization of richer and richer models of grammar.
